---
title: "Calibration metadata"
author: "Irene Solano"
format: html
editor: visual
execute:
  echo: true
  warning: false
  message: false
---

# Description

The **NaAMPHoRA database** includes modern and fossil pollen records along with their chronologies. Raw radiocarbon dates were calibrated using a Bayesian age-depth modeling approach implemented in the ***rbacon*** **R package** (v3.2.0; Blaauw & Christen, 2011). For sites with only one dated depth or missing depth information, we calibrated radiocarbon ages using the ***rice*** **R package** (v1.0.0; Blaauw, 2024).

Details on the calibration for each site are available in the *calibration_information.csv* file, which contains the following metadata:

```{r}
#| echo: false
#| output: false
#| message: false
# Load libraries and functions ----
source("functions.R")

library <- c("knitr", "dplyr","readr")

# Install missing packages
invisible(lapply(library, install_if_missing))

# Load the library
lapply(library, require, character.only = TRUE)

```

```{r}
#| echo: false
#| output: false
#| message: false
# Define Attributes

# Define variable names, descriptions, and units

# View the dataframe
attributes <- data.frame(
  Name = c("Biogeographic area", "Country", "Site", "Site_name_machine_readable", 
           "Dated", "Recalibrated", "Calibration R package", "Radiocarbon dates", 
           "Depths", "Material dated", "Dating method (Conventional/AMS)", 
           "Calibration curve", "Hiatus", "Post-bomb curve", 
           "Reservoir effect", "Accumulation rate (yr/cm)", "Section thickness (cm)", 
           "Comments #1", "Comment #2"),
  Description = c("Biogeographical region to which the site belongs", 
                  "Country where the site is located", "Site name", 
                  "Machine-readable name for the site", "Whether the record was dated", 
                  "Whether the record was recalibrated in this study", "R package used for calibration", 
                  "Whether the record had raw radiocarbon dates with its associated error", "Whether the record reports depths along with the radicarbon dates", 
                  "Material used for dating (e.g., wood, sediment)", 
                  "Method used for radiocarbon dating (Conventional or AMS)", 
                  "Calibration curve used in the calibration process", 
                  "Whether there was a hiatus in the record", 
                  "Whether there was a post-bomb dated and the corresponding Post-bomb curve used in calibration", 
                  "Whether there was an effect of reservoir on radiocarbon dating", 
                  "Rate of accumulation in years per cm used in the calibration", "Thickness of the section in cm used in the calibration", 
                  "Additional comments regarding the site or data", 
                  "Further comments or notes on the dataset"),
  Unit = c("Categorical", "Categorical", "Categorical", "Categorical", "Categorical", "Categorical", "Categorical", "Categorical", "Categorical", "Categorical", "Categorical","Categorical", "Categorical", "Categorical", "Categorical", "Years/cm", "Centimeters","Categorical", "Categorical")
)



# Save attributes metadata as CSV
write_csv(attributes, normalizePath("../metadata/age_calibrated/calibration_metadata.csv"))

```

# Metadata

```{r}
#| echo: false
# Display table in HTML
kable(attributes)
```

# References

1.  Blaauw, M., & Christen, J. A. Flexible paleoclimate age-depth models using an autoregressive gamma process. *Bayesian Analysis, 6*(3), 457–474. <https://doi.org/10.1214/11-BA618> (2011).

2.  Blaauw, M. *rice: Radiocarbon Equations*. <https://CRAN.R-project.org/package=rice> (2024).
